---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

<!-- badges: start -->
<!-- badges: end -->

The goal of fpod is to provide a means to directly load FPOD and CPOD data files into R. The FPOD data files contain binary data, so they can’t trivially be read into R using the usual approach (e.g. fread or read.csv). This package decodes the binary data and imports all the data in one go (i.e. header/metadata, clicks, KERNO classifications, environmental data and pseudo-WAV data). It is then trivial to aggregate data as you please, e.g. DPMs per time block. The advantage of handling data processing in R is a long topic, but suffice it to say that it 1) simplifies things (many fewer steps, as different vars have to be exported in multiple goes in the official FPOD app), and more importantly, 2) makes data processing 100% transparent and reproducible.

## Overview
* `fp_read()` reads a CP1, CP3, FP1 or FP3 binary data file
* `fp_summarize()` calculates minute-resolution summaries of clicks
* `fp_find_buzzes()` finds harbour porpoise buzzes in a subset of clicks

## Installation

You can install the development version of fpod from [GitHub](https://github.com/) with:

``` r
#install.packages("devtools")
library(devtools)
install_github("supermoan/fpod")
```

## Quick start
If you just want to get started, load the package, and call fp_read on your FP3 file. For example:
```{r example}
library(fpod)
fn <- fp_example("gullars_period1.FP3")
dat <- fp_read(fn)
```
This will give you access to most of the information in the FP3 file, including header/metadata, clicks data, wav, data and misc other data, such as battery levels. 

For more examples, see the vignette and the help files for each individual function.
